GVN-Hoist: Hoisting Computations from Branches

### 2016-08-23 hoist.tex
more cleanups

\subsection{Finding the candidates}
The first step is to find a set of instructions that perform identical
computations: this is performed by a linear scan of all instructions of the
program and classifying all instructions by their value given by GVN.

@@

The process of computing GVN can be on-demand (as we come across an instruction)
or, precomputed (computing GVN of all the instructions beforehand). Which
process to choose is determined by the scope of code-hoisting we want to
perform. In a pessimistic approach, as described in
Section-\ref{subsec:pessimistic}, we want to hoist a limited set of instructions
from the sibling branches as we iterate the DFS tree bottom-up, it is sufficient
to compute GVN values on-demand. Whereas, in the optimistic approach, as
described in Section-\ref{subsec:optimistic}, we want to hoist as many
instructions as possible, and it would require GVN values to be precomputed.

@@

\subsection{Checking for legality}
\label{subsec:legality}
Since the equality of candidates is purely based on the value they compute, we
need to establish if hoisting them to a common dominator would be feasible. Once
a common dominator is found, we check whether all the use-operands of the set of
instructions are available at that position. In some cases when the operands are not
available, it is possible to reinstantiate (remateralize) the use-operands, thus
passing the legality check.

@@

In the optimistic approach, described in Section-\ref{subsec:optimistic}, it is
possible that a common hoisting point of all the instructions is either too far
away, or not legally possible. In these cases, it is still possible to
`partially' hoist a subset of instructions by splitting the set of candidates
and finding a closer hoisting point for each subset. For more details see
Section-\ref{subsec:partition}.

\subsubsection{Hoisting scalars}
Scalars are the easiest to hoist because we do not have to analyze them for
aliasing memory references. As long as all the operands are available (or can be
made available by rematerialization), the scalar computation can be hoisted.

\subsubsection{Hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the addresss if
possible.  Along the path, from current position of the load instruction
backwards on the control flow to the hoisting point, we check whether there are
writes to memory that may alias with the load, in which case we discard the
candidate.

\subsubsection{Hoisting stores}
For stores, we check the dependency requirements similar to the hoisting of
loads. We check that the operands of the store instruction are available at the
hoisting point, that there are no aliasing loads or store along the path from
the current position to the hoisting point.

\subsubsection{Hoisting calls}
Call instructions can be divided into three categories: those calls equivalent
to purely scalar computations, calls reading from memory, and most of the time,
without further information, calls have to be classified as writing to memory,
that is the most restrictive form.  Each category of call instructions will be
handled as described for scalar, load, and store instructions.

\subsection{Checking for profitability}
\label{subsec:profitability}
After the legality checks have passed, we check for profitability of hoisting.
That takes into account the impact code-hoisting would have on various
parameters that affects runtime performance e.g., impact on live-range, gain in
the code size.  We have established a set of cost models \ref{sec:cost-models}
for each parameter and tuned them for performance against representative
benchmarks.

\subsubsection{Hoisting scalars}
Since scalars are the majority of instructions which are hoisted, we pay special
attention in case of hoisting scalars too far, as that may increase register
pressure and result in spills. For example hoisting a scalar past a call
(Section-\ref{cost:across-calls}).  In our current implementation we hoist
scalars past a call only when optimizing for code-side (-Os). Ideally, a later
stage of live-range splitting pass should split the live-ranges for optimal
performance, however, that is not the case with llvm as we have found
regressions when scalars are hoisted too far \ref{sec:cost-models}. Another way
to mitigate this problem is be to reinstantiate (rematerialize) the computation
after a call (may be as a different optimization pass).

\subsubsection{Hoisting loads}
A load instruction introduces a register where the value loaded will be kept,
the register pressure increases by one (unless the operand to load becomes dead
at the load). On the other hand, loading a value early will reduce the stall
during execution should the value is not in the cache. We generally prefer to
hoist load except the hoisting point is too far (this distance is computed by
looking at the experimental results of representative benchmarks see
Section-\ref{sec:experimental-results}).

\subsubsection{Hoisting stores}
Since stores do not increase the live-range of any registers, and in some cases
it ends the liveness of registers, we hoist all the stores.

\subsubsection{Hoisting calls}
Currently we hoist all the calls that are suitable candidates for hoisting.

\subsection{Making the transformation}
Once all the legality checks and profitability checks are satisfied for a set of
identical instructions, they are suitable candidates for hoisting. A copy of the
computation is inserted at the hoisting point along with any instructions which
needed to be rematerialized. Thereafter, all the computations made redundant by
the new copy are removed. After that SSA form is restored by updating the
intermediate representation (IR) to reflect changes. After one iteration of
algorithm runs through the entire function, it creates more oppportunities for
\emph{higher ranked} computations \cite{rosen1988global}. Currently, we rerun
the algorithm until there are no more instructions left to be
hoisted. Obviously, this is not the most optimal approach and can be improved by
ranking the computations \cite{rosen1988global}.

@@

\subsection{Hoisting bottom up with optimistic approach}
\label{subsec:optimistic}
In this approach, the goal is to maximise the total number of hoistings in the
entire function.  This algorithm is very useful when optimizing for code-size.
We collect the GVN of all the instructions in the function and iterate on the
list of instructions having identical GVNs. After that we find the common
dominator dominating all such identical computations and perform legality checks
(Section-\ref{subsec:legality}). Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints e.g.,
intersecting side-effects or, profitability constraints e.g., hoisting point too
far. In those cases, this algorithmn would partition the list of identical
instructions into subsets which can be partially hoisted to their respective
common dominators. The partition algorithm is described as follows:

@@

In our current implementation we keep as many candidates in one set as possible
(greedy approach). We split the list at a point where the legality checks fail
to hoist subset of candidates which are legal to hoist and then start finding
new hoisting point for the remaining ones. So there are two limitations of the
current implementation of the partition algorithm: First, sorting by DFSIn
numbers does not give the desired order when there are fully redundant
instructions; Second, it lacks cost model for partitioning in order to
preserve/improve performance \ref{sec:future-work}.

@@

\subsection{In the presence of calls}
\label{cost:across-calls}
Hoisting scalars across calls is tricky because it can increase the number of
spills. During the frame lowering of calls, the argument registers, in general,
the caller saved registers are saved because they might be modified by the
callee and after the call they are restored \cite{frame-lowering}. So before the
call, the register pressure is high because the number of available registers
are reduced by the number of caller saved registers. In that situation if a
computation is hoisted across the call, that would increase the total number of
registers required by one, thus contributing to the register pressure. However,
in the special case discussed in Section-\ref{hoist:reg-pressure}, it will be
okay to hoist because the register pressure would not decrease.

@@

However, in the special case discussed in Section-\ref{hoist:reg-pressure}, it
will be okay to hoist because the register pressure would not decrease.

\section{Experimental Evaluation}
\label{sec:experimental-results}
We ran llvm-testsuite (trunk:d87471f8) with the patch (trunk:86940146) and the
results are listed in Table-\ref{tab:hoist-results}. The table lists the number
of scalars, loads, stores and calls hoisted as well as removed. For each
category, the number of instructions removed is greater or equal to the number
of instructions hoisted because each hoisting is performed only when at least
one identical computation is found.

@@

Other static metrics are listed in Table-\ref{tab:static-results}. Here we can
see that except for rematerializing defs for splitting, which has an overhead of ~2\%, all
other parameters have less than 1\% overhead. This is to explain why the performance does not
go down with our implementation (and cost-model) of code hoisting pass.

@@

While benchmarking llvm-testsuite we see both increase as well as decrease in
the codesizes of the final binaries. Since the pass runs early, it has affect on
many optimizations which rely on number of instructions, lenth of the use-def
chain etc metrics which are affected by code-hoisting. Specially the inliner
where the total inline cost (in llvm) is heavily dependent on the total count of
instructions in the caller and callee. Various code-size metrics are shown in
Table-\ref{tab:code-size}. All but one benchmark varied between -5.32\% and
5.43\%.  In one benchmark FreeBench/distray/distray.test, the codesize increased
by 35.38\%. In this benchmark 3 more (15 as compared to 12) functions got
inlined and because of that 10 more (81 vs. 71) vector instructions got
generated, 3 calls got hoisted/sunk as (compared to 0), one loop got unswitched
(compared to 0), 6 high latency machine instructions got hoisted out of loop, 59
(compared ot 30) machine instructions got hoisted out of loop, 70 (compared to
39) machine instructions were sunk.

\section{Conclusion and Future Work}
\label{sec:future-work}
We have presented the GVN based code hoisting algorithm. The primary goal is to
reduce the code size but it benefits performance in some cases as well. To
preserve performance and not hoist too much we have implemented several cost
models \ref{sec:cost-models}. Since those cost models depend on a set of thresholds,
it requires tuning, as such, we used representative benchmarks to tune them.

@@

With the implementation of code-hoisting in llvm, the passes which rely on the
code-size/instruction-count to make optimization decisions needs to be
revisited. The first candidate would be the inliner. We have seen different
inlining decisions (See Table-\ref{tab:code-size}) before and after
code-hoisting was enabled.  Since inliner has several magic numbers tuned for
the previous pass layout, it would need some improvement.

>>>>

\subsection{Finding candidates to hoist}
The first step is to find a set of instructions that perform identical
computations: this is performed by a linear scan of all instructions of the
program and classifying all instructions by their value given by GVN.

@@

The process of computing GVN can be on-demand (as we come across an instruction)
or, precomputed (computing GVN of all the instructions beforehand). Which
process to choose is determined by the scope of code-hoisting we want to
perform. In a pessimistic approach, as described in
Section~\ref{subsec:pessimistic}, we want to hoist a limited set of instructions
from the sibling branches as we iterate the DFS tree bottom-up, it is sufficient
to compute GVN values on-demand. Whereas, in the optimistic approach, as
described in Section~\ref{subsec:optimistic}, we want to hoist as many
instructions as possible, and it would require GVN values to be precomputed.

@@

\subsection{Legality check}
\label{subsec:legality}
Since the equality of candidates is purely based on the value they compute, we
need to establish if hoisting them to a common dominator would be feasible. Once
a common dominator is found, we check whether all the use-operands of the set of
instructions are available at that position. In some cases when the operands are not
available, it is possible to reinstantiate (remateralize) the use-operands, thus
passing the legality check.

@@

In the optimistic approach, described in Section~\ref{subsec:optimistic}, it is
possible that a common hoisting point of all the instructions is either too far
away, or not legally possible. In these cases, it is still possible to
`partially' hoist a subset of instructions by splitting the set of candidates
and finding a closer hoisting point for each subset. For more details see
Section~\ref{subsec:partition}.

\subsubsection{Legality of hoisting scalars}
Scalars are the easiest to hoist because we do not have to analyze them for
aliasing memory references. As long as all the operands are available (or can be
made available by rematerialization), the scalar computation can be hoisted.

\subsubsection{Legality of hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the addresss if
possible.  Along the path, from current position of the load instruction
backwards on the control flow to the hoisting point, we check whether there are
writes to memory that may alias with the load, in which case we discard the
candidate.

\subsubsection{Legality of hoisting stores}
For stores, we check the dependency requirements similar to the hoisting of
loads. We check that the operands of the store instruction are available at the
hoisting point, that there are no aliasing loads or store along the path from
the current position to the hoisting point.

\subsubsection{Legality of hoisting calls}
Call instructions can be divided into three categories: those calls equivalent
to purely scalar computations, calls reading from memory, and most of the time,
without further information, calls have to be classified as writing to memory,
that is the most restrictive form.  Each category of call instructions will be
handled as described for scalar, load, and store instructions.

\subsection{Profitability check}
\label{subsec:profitability}
After the legality checks have passed, we check for profitability of hoisting.
That takes into account the impact code-hoisting would have on various
parameters that affect runtime performance e.g., impact on live-range, gain in
the code size.  We have established a set of cost models described in Section~\ref{sec:cost-models}
for each parameter and tuned them for performance against representative
benchmarks.

\subsubsection{Profitability of hoisting scalars}
Since scalars are the majority of instructions which are hoisted, we pay special
attention in case of hoisting scalars too far, as that may increase register
pressure and result in spills. For example hoisting a scalar past a call, as
described in Section~\ref{cost:across-calls}.  In our current implementation we
hoist scalars past a call only when optimizing for code-side (-Os). Ideally, a
later stage of live-range splitting pass should split the live-ranges for
optimal performance, however, that is not the case with llvm as we have found
regressions when scalars are hoisted too far, as in
Section~\ref{sec:cost-models}. Another way to mitigate this problem is be to
reinstantiate (rematerialize) the computation after a call (may be as a
different optimization pass).

\subsubsection{Profitability of hoisting loads}
A load instruction introduces a register where the value loaded will be kept,
the register pressure increases by one (unless the operand to load becomes dead
at the load). On the other hand, loading a value early will reduce the stall
during execution should the value is not in the cache. We generally prefer to
hoist load except the hoisting point is too far (this distance is computed by
looking at the experimental results of representative benchmarks, see
Section~\ref{sec:experimental-results}).

\subsubsection{Profitability of hoisting stores}
Since stores do not increase the live-range of any registers, and in some cases
it ends the liveness of registers, we hoist all the stores.

\subsubsection{Profitability of hoisting calls}
Currently we hoist all the calls that are suitable candidates for hoisting.

\subsection{Code generation}
Once all the legality checks and profitability checks are satisfied for a set of
identical instructions, they are suitable candidates for hoisting. A copy of the
computation is inserted at the hoisting point along with any instructions which
needed to be rematerialized. Thereafter, all the computations made redundant by
the new copy are removed. After that SSA form is restored by updating the
intermediate representation (IR) to reflect changes. After one iteration of
algorithm runs through the entire function, it creates more oppportunities for
\emph{higher ranked} computations \cite{rosen1988global}. Currently, we rerun
the algorithm until there are no more instructions left to be
hoisted. Obviously, this is not the most optimal approach and can be improved by
ranking the computations \cite{rosen1988global}.

@@

\subsection{Hoisting bottom up with optimistic approach}
\label{subsec:optimistic}
In this approach, the goal is to maximise the total number of hoistings in the
entire function.  This algorithm is very useful when optimizing for code-size.
We collect the GVN of all the instructions in the function and iterate on the
list of instructions having identical GVNs. After that we find the common
dominator dominating all such identical computations and perform legality
checks, as described in Section~\ref{subsec:legality}. Often times it is not
possible to hoist all the instructions to one common dominator, due to legality
constraints e.g., intersecting side-effects or, profitability constraints e.g.,
hoisting point too far. In those cases, this algorithmn would partition the list
of identical instructions into subsets which can be partially hoisted to their
respective common dominators. The partition algorithm is described as follows:

@@

In our current implementation we keep as many candidates in one set as possible
(greedy approach). We split the list at a point where the legality checks fail
to hoist subset of candidates which are legal to hoist and then start finding
new hoisting point for the remaining ones. So there are two limitations of the
current implementation of the partition algorithm: First, sorting by DFSIn
numbers does not give the desired order when there are fully redundant
instructions; Second, it lacks cost model for partitioning in order to
preserve/improve performance, see Section~\ref{sec:future-work}.

@@

\subsection{In the presence of calls}
\label{cost:across-calls}
Hoisting scalars across calls is tricky because it can increase the number of
spills. During the frame lowering of calls, the argument registers, in general,
the caller saved registers are saved because they might be modified by the
callee and after the call they are restored \cite{frame-lowering}. So before the
call, the register pressure is high because the number of available registers
are reduced by the number of caller saved registers. In that situation if a
computation is hoisted across the call, that would increase the total number of
registers required by one, thus contributing to the register pressure. However,
in the special case discussed in Section~\ref{hoist:reg-pressure}, it will be
okay to hoist because the register pressure would not decrease.

@@

However, in the special case discussed in Section~\ref{hoist:reg-pressure}, it
will be okay to hoist because the register pressure would not decrease.

\section{Experimental Evaluation}
\label{sec:experimental-results}
We ran llvm-testsuite (trunk:d87471f8) with the patch (trunk:86940146) and the
results are listed in Table~\ref{tab:hoist-results}. The table lists the number
of scalars, loads, stores and calls hoisted as well as removed. For each
category, the number of instructions removed is greater or equal to the number
of instructions hoisted because each hoisting is performed only when at least
one identical computation is found.

@@

Other static metrics are listed in Table~\ref{tab:static-results}. Here we can
see that except for rematerializing defs for splitting, which has an overhead of ~2\%, all
other parameters have less than 1\% overhead. This is to explain why the performance does not
go down with our implementation (and cost-model) of code hoisting pass.

@@

While benchmarking llvm-testsuite we see both increase as well as decrease in
the codesizes of the final binaries. Since the pass runs early, it has affect on
many optimizations which rely on number of instructions, lenth of the use-def
chain etc metrics which are affected by code-hoisting. Specially the inliner
where the total inline cost (in llvm) is heavily dependent on the total count of
instructions in the caller and callee. Various code-size metrics are shown in
Table~\ref{tab:code-size}. All but one benchmark varied between -5.32\% and
5.43\%.  In one benchmark FreeBench/distray/distray.test, the codesize increased
by 35.38\%. In this benchmark 3 more (15 as compared to 12) functions got
inlined and because of that 10 more (81 vs. 71) vector instructions got
generated, 3 calls got hoisted/sunk as (compared to 0), one loop got unswitched
(compared to 0), 6 high latency machine instructions got hoisted out of loop, 59
(compared ot 30) machine instructions got hoisted out of loop, 70 (compared to
39) machine instructions were sunk.

\section{Conclusion and Future Work}
\label{sec:future-work}
We have presented the GVN based code hoisting algorithm. The primary goal is to
reduce the code size but it benefits performance in some cases as well. To
preserve performance and not hoist too much we have implemented several cost
models described in Section~\ref{sec:cost-models}. Since those cost models
depend on a set of thresholds, it requires tuning, as such, we used
representative benchmarks to tune them.

@@

With the implementation of code-hoisting in llvm, the passes which rely on the
code-size/instruction-count to make optimization decisions needs to be
revisited. The first candidate would be the inliner. We have seen different
inlining decisions in Table~\ref{tab:code-size}, before and after
code-hoisting was enabled.  Since inliner has several magic numbers tuned for
the previous pass layout, it would need some improvement.
